---
title: Geographic data in R
format: 
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
---

```{r}
#| label: load-packages
library(sf)
library(spData)
library(tidyverse)
```

# Prerrequisites

Install the following packages:


```r
install.packages("sf")
install.packages("terra")
install.packages("spData")
install.packages(
    "spDataLarge", 
    repos = "https://geocompr.r-universe.dev")
```

For linux check the [Posit Public Package Manager](https://packagemanager.posit.co) and apply the `SETUP` for your case. This repository is configured by default using Positron. Check:

```r
getOption("repos")
```

For example for `Ubuntu 24.04 LTS`:

```
                                                          P3M 
"https://packagemanager.posit.co/cran/__linux__/noble/latest" 
                                                         CRAN 
                                "https://cloud.r-project.org" 
```

# Introduction

Fundamental geographic data models:

- Vector
- Raster

The **vector data model** represents the world using points, lines and polygons. The **raster data model** divides the surface up into cells of constant size.

Which to use?

- Vector data tends to dominate the social sciences because human settlements tend to have discrete borders
- Raster dominates many enfironmental sciences partially becasue the reliance on remote sensing data

# Vector data

The geographic vecto data model is based on points located withing a coordinate reference system, **CRS**.

Points ca represent self-standing features or they can be linked together to form more complex geometries such as lines and polygons.

IN this system, for example, London can be represented by the coordinates `c(-0.1, 51.5)`. This means that its location is `-0.1` degrees east and `51.5` degrees north of the origin. The origin in this case is at `0` degrees longitude (Prime Meridian) and `0` degrees latitude (Equator) in a geographic (lon/lat) CRS.

For this section, it is sufficient to know that coordinates consist of two numbers representing distances from an origin, usually in $x$ and $y$ dimensions.

## `sf` package

- **GDAL**: reading, writing and manipulating geographic data formats
- **PROJ**: coordinate system transformations
- **GEOS**: planar geometry engine for operations such as calculating buffers and centroids on data with a projected CRS
- **S2**: a spherical geometry engine

As an example when loading `sf`:

```
Linking to GEOS 3.12.1, GDAL 3.8.4, PROJ 9.4.0; sf_use_s2() is TRUE
```

### Simple features

Simple features is an open standard developed and endorsed by the Open Geospatial Consortium (OGC). Simple features is a hierarchical data model taht represents a wide rande of geometry types.

Simple feature types fully supported by `sf`:

- Multilinestring
- Multipoint
- Multipolygon
- Geometry collection
- Polygon
- Point
- Line string

Raster data classes are not supported by `sf`. For more information check the site [Simple Features for R](https://r-spatial.github.io/sf/) and the vignettes:

```r
vignette(package = "sf")
vignetter("sf1")
```

Simple feature objects in R are store in a data frame, with geographic data occupying a special column, usually name `geom` or `geometry`.

We are going to use the `world` dataset provided by `spData`:

```{r}
class(world)
names(world)
```

`sf` objects can be plotted quickly with the function `plot()`. For example:

```{r}
plot(world, max.plot = 10)
```

In this case 10 variables from `world` are plot except `geom` where are the polygons. Also treating geographic objects as regular data frames with spatial powers nas many advantages, especially if you are already used to working with data frames.

```{r}
world["lifeExp"] |>
  summary()
```

In this case also the geometry (`geom`) is reported and the world `MULTIPOLYGON` in the output referts to the geometry type of features (countries) in the `world` object.

`sf` objects are easy to subset. For example you can return only the first 2 rows and the first 3 columns of the `world` object.

```{r}
world_mini <- world[1:2, 1:3]
world_mini
```

In a `sf` object, the geometry column is considered the defining characteristic of the data. Without the geometry, the object ceases to be spatial. That is why, `sf` is designed to ensure that the spatial integrity of the features remains intact.

The output shows two major differences compared with a regular `data.frame`: the inclusion of additional geographic metadata (`Geometry type`, `Dimension`, `Bounding box` and coordinate references system information). Also it has the `geom` column.

The `geom` column are list columns of class `sfg`:

```{r}
world$geom |>
  class()
```

`sfc` objects are composed of one or more objects of class `sfg`: simple features geometries.

### Why simple features?

- Fast reading and writing of data
- Enhanced plotting performance
- `sf` objects can be treated as data frames in most operations
- `sf` function names are relatively consistent and intuitive (all begin with `st_`)
- `sf` functions can be combined with the `|>` operator and works well with the `tidyverse` collection of R packages

`sf` support for the `tidyverse` packages is exemplified by `read_sf()`, a function for importing geographic vector data. Unlike the function `st_read()`, which returns attributes stored in a base R `data.frame`, `read_sf()` silently returns data as a tidyverse `tibble`. For example:

```{r}
world_dfr <- system.file(
  "shapes/world.gpkg",
  package = "spData"
) |>
  st_read()

world_tbl <- system.file(
  "shapes/world.gpkg",
  package = "spData"
) |>
  read_sf()

class(world_dfr)
class(world_tbl)
```

`world_tbl` include also the classes `tbl_df` and `tbl` additionally to `sf` and `data.frame`.

### Basic maps

Basic geographic visualizations (maps) are created in `sf` with base R `plot()` function. Be default, this creates a multi-panel plot, one sub-plot for each variable of the object. For example:

```{r}
plot(world[3:6])
plot(world["pop"])
```

`plot()` is a generic method that behaves differently depending on the class of object beign plotted. For example, you can check `methods(plot)`:

```{r}
methods(plot)[methods(plot) == "plot.sf"]
```

Plots are added as layers to existing images by setting `add = TRUE`. For example lets filter the countries in Asia and combines then into a single feature:

```{r}
world_asia <- world[world$continent == "Asia", ]
# A single MULTIPOLYGON is created that represents Asia
# as a single unit
asia <- st_union(world_asia)
plot(asia)
```

We can now plot the Asian continent over a map of the world. The first plot must only have one facet for `add = TRUE`. Also if the first plot has a key (a legend, for example a color scale to represent the data) then you need to specify `reset = FALSE`

```{r}
plot(world["pop"], reset = FALSE)
plot(asia, add = TRUE, col = "red")
```

**Note**: This plot is problematic because the legend shows a fill scale to represent the population for the countries except asia.

Lets try another visualization by adding circles that represent the population on a map of the world as an unprojected version:

```{r}
plot(world["continent"], reset = FALSE)
cex <- sqrt(world$pop) / 10000
world_cents <- st_centroid(world, of_largest_polygon = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)
```

In this case `st_centroid` gives the centroid of a geometry and the argument `of_largest_polygon` with the option `TRUE` finds the single part with the greatest area and uses its center. This is because many countries are multipoligons so by default a centroid might end up in the ocean between the mainland and an island.

`sf` plot method also has arguments specific to geographic data, `expandBB`, for example, can be used to plot and `sf` object in context: it takes a numeric vector of length four that expands the bounding box of the plot relative to zero in the following order:

- Bottom
- Left
- Top
- Right

This is used to plot india in the context of its giant Asian neighbors, with and emphasis on China to the east, in the following code chunk:

```{r}
india = world[world$name_long == "India", ]
plot(
  st_geometry(india),
  expandBB = c(0, 0.2, 0.1, 1),
  col = "gray",
  lwd = 3
)
plot(st_geometry(world_asia), add = TRUE)
```

### Geometry types

Geometries are the basic building blocks of simple features. Simple features in R can take on one of the 18 geometry types supported by the `sf` package. In this part we will focus on:

- `POINT`
- `LINESTRING`
- `POLYGON`
- `MULTIPOINT`
- `MULTILINESTRING`
- `MULTYPOLYGON`
- `GEOMETRYCOLLECTION`

The basis for each geometry type is the point. A point is simply a coordinate in two-, three or four-dimensional space (see `vignetter("sf1")` for more information):

- `POINT (5,2)`

A linestring is a sequence of points with a straight line connectiong the points:

- `LINESTRING (1 5, 4 4, 4 1, 2 2, 3 2)`

A polygon is a sequence of points that form a closed, non-intersecting ring. Closed means that the first and the last point of a polygon have the same coordinates

- `POLYGON ((1 5, 2 2, 4 1, 4 4, 1 5))`

**Note**: See `1 5` at the start and end 

So far we have created geometries with only one geometric entity per feature. Simple feature standard also allows multiple geometries of a single type to exist within a single feature with "multi" version of each geometry type.

- Multipoint: `MULTIPOINT (5 2, 1 3, 3 4, 3 2)`
- Multilinestring: `MULTILINESTRING ((1 5, 4 4, 4 1, 2 2, 3 2), (1 2, 2 4))`
- Multipolygon: `MULTIPOLYGON ((1 5, 2 2, 4 1, 4 4, 1 5), (0 2, 1 2, 1 3, 0 3, 0 2))`

Finally a geometry collection can contain any combination of geometries including (multi)points and linestrings.

- Geometry collection: `GEOMETRYCOLLECTION (MULTIPOINT (5 2, 1 3, 3 4, 3 2), LINESTRING (1 5, 4 4, 4 1, 2 2, 3 2))`

### The sf class

Simple features consist of 2 main parts:

- Geometries
- Non-geographic attributes

Geometries come from a `sfc` object while non-geographic attributes are taken from a `data.frame` or `tibble`. Non-geographic attributes represent the name of the feature or other attributes such as measured values, groups and other things.

To illustrate attribures, we will represent a temperature of 25°C in London on June 21, 2023. This example contains a geometry (coordinates) and 3 attributes with 3 different classes (place name, temperature and date). Objects of class `sf` represent such data by combining the attributes (`data.frame`) with simple feature geometry column (`sfc`). They are created with `st_sf()` with the simple feature geometry column (`sfc`). They are created with `st_sf()` as illustrated below, which creates the London example described above:


```{r}
lnd_point = st_point(c(0.1, 51.5)) # sfg object
lnd_geom = st_sfc(lnd_point, crs = "EPSG:4326") # sfc object
lnd_attrib = tibble(
  name = "London",
  temperature = 25,
  date = ymd("2023-06-21")
)
lnd_sf = st_sf(lnd_attrib, geometry = lnd_geom)
```

What just happened? First, the coordinates were used to create the simple feature geometry (`sfg`). Second, the geometry was converted into a simple feature geometry column (`sfc`), with a CRS. Third, attributes were stored in a `tibble`, which has combined with the `sfc`, with `st_sf()`. This results in an `sf` object, as demonstrated below:

```{r}
lnd_sf
class(lnd_point)
class(lnd_geom)
class(lnd_sf)
```

### Simple feature geometries (`sfg`)

The `sfg` class represents the different simple feature geomtry types in R:

- Point
- Linestring
- Polygon

And their "multi" equivalents.

Usually you are spared the tedious task of creating geometries on your own since you can simply import and already existing spatial file. However, there are a set of functions to create simple feature geomtry objects (`sfg`) from scratch, if needed. 

The names of these functions are simple and consistent, as the all start with `st_` prefix and end with the name of the geometry type:

- A point: `st_point`
- A linestring: `st_linestring()`
- A polygon: `st_polygon`
- A multipoint: `st_multipoint()`
- A multilinestring: `st_multiline_string()`
- A multipolygon: `st_multipolygon()`
- A geometry collection: `st_geometrycollection()`

`sfg` objects can be created from 3 base R data types:

1. A numeric vector: a single point
2. A matrix: a set of points, where each row represents a point, a multipoint or linestring
3. A list: a colletion of objects such as matrices, multilinestrings or geometry collections

The function `st_point()` creates single points form numeric vectors:

```{r}
st_point(c(5, 2)) # XY point (2D)
st_point(c(5, 2, 3)) # XYZ point (3D)
st_point(c(5, 2, 1), dim = "XYM") # XYM point (3D with an additional variable, typically measurement accuracy)
st_point(c(5, 2, 3, 1)) # XYZM
```

The `XYM` type must be specified using the `dim` argument (which is a short for dimension)

By contrast, use matrices in the case of multipoint (`st_multipoint()`) and linestring (`st_linestring`) objects:

```{r}
multipoint_matrix = rbind(
  c(5, 2),
  c(1, 3),
  c(3, 4),
  c(3, 2)
)
st_multipoint(multipoint_matrix)

linestring_matrix = rbind(
  c(1, 5),
  c(4, 4),
  c(4, 1),
  c(2, 2),
  c(3, 2)
)
st_linestring(linestring_matrix)
```

Finally, use lists for the creation of multilinestrings, (multi-)polygons and geometry collections:

```{r}
## POLYGON
polygon_lists = list(
  rbind(
    c(1, 5),
    c(2, 2),
    c(4, 1),
    c(4, 4),
    c(1, 5)
  )
)
st_polygon(polygon_lists)
```


```{r}
# POLYGON with a hole
polygon_border = rbind(
  c(1, 5),
  c(2, 2),
  c(4, 1),
  c(4, 4),
  c(1, 5)
)
polygon_hole = rbind(
  c(2, 4),
  c(3, 4),
  c(3, 3),
  c(2, 3),
  c(2, 4)
)
polygon_with_hole_list = list(
  polygon_border,
  polygon_hole
)
st_polygon(polygon_with_hole_list)
```

```{r}
# MULTILINESTRING
multilinestring_list = list(
  rbind(
    c(1, 5),
    c(4, 4),
    c(2, 2),
    c(3, 2)
  )
)
st_multilinestring(multilinestring_list)
```

```{r}
# MULTIPOLYGON
multipolygon_list = list(
  list(
    rbind(
      c(1, 5),
      c(2, 2),
      c(4, 1),
      c(4, 4),
      c(1, 5)
    )
  ),
  list(
    rbind(
      c(0, 2),
      c(1, 2),
      c(1, 3),
      c(0, 3),
      c(0, 2)
    )
  )
)
st_multipolygon(multipolygon_list)
```

```{r}
## GEOMETRYCOLLECTION
geometrycollection_list = list(
  st_multipoint(multipoint_matrix),
  st_linestring(linestring_matrix)
)
st_geometrycollection(geometrycollection_list)
```

### Simple feature columns (sfc)

One `sfg` object contains only a single simple feature geometry. A simple feature geometry column (`sfc`) is a list of `sfg` objects, which is additionaly abel to contain information about (`sfc`) is a list of `sfg` objects, which is additionally able to contain information about the Coordinate Reference System - CRS in use.4

For instance, to combine 2 simploe features into one object with 2 features, we can use the `st_sfc()` function. This is important since `sfc` represents the geometry column in `sf` data frames:

```{r}
# sfc POINT
point1 = st_point(c(5, 2))
point2 = st_point(c(1, 3))
points_sfc = st_sfc(point1, point2)
points_sfc
```

In most cases, an `sfc` object contains objects of the same geometry type. Therefore, when we convert `sfg` objects of type plygon into a simple feature geometry column, we would also end up with an `sfc` object of type polygon, which can be verified with `st_geometry_type()`. Equally, a geometry column of multilinestrings would result in an `sfc` object of type multilinestring:

```{r}
# sfc POLYGON
polygon_list1 = list(
  rbind(
    c(1, 5),
    c(2, 2),
    c(4, 1),
    c(4, 4),
    c(1, 5)
  )
)
polygon1 = st_polygon(polygon_list1)
polygon_list2 = list(
  rbind(
    c(0, 2),
    c(1, 2),
    c(1, 3),
    c(0, 3),
    c(0, 2)
  )
)
polygon2 = st_polygon(polygon_list2)
polygon_sfc = st_sfc(
  polygon1,
  polygon2
)
st_geometry_type(polygon_sfc)
```

```{r}
multilinestring_list1 = list(
  rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2)),
  rbind(c(1, 2), c(2, 4))
)
multilinestring1 = st_multilinestring((multilinestring_list1))
multilinestring_list2 = list(
  rbind(c(2, 9), c(7, 9), c(5, 6), c(4, 7), c(2, 7)),
  rbind(c(1, 7), c(3, 8))
)
multilinestring2 = st_multilinestring((multilinestring_list2))
multilinestring_sfc = st_sfc(multilinestring1, multilinestring2)
st_geometry_type(multilinestring_sfc)
```

It is also possible to create an `sfc` object from `sfg` objects with different geometry types:

```{r}
point_multilinestring_sfc = st_sfc(
  point1,
  multilinestring1
)
st_geometry(point_multilinestring_sfc)
```

As mentioned before, `sfc` objects can additionally store information on the CRS. The default values is `NA` (NOT AVAILABLE), as can be verified with `st_crs()`:

```{r}
st_crs(points_sfc)
```

All geometries in `sfc` objects must have the same CRS. ACRS can be specifies with the `crs` argument of `st_sfc()` (or `st_sf()`), which takes a CRS identifier provied as a text string, such as `crs = "EPSG:4326"`


```{r}
points_sfc_wgs = st_sfc(
  point1,
  point2,
  crs = "EPSG:4326"
)
st_crs(points_sfc_wgs)
```

## The sfheaders package