---
title: Geographic data in R
format: 
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
---

```{r}
#| label: load-packages
library(sf)
library(spData)
```

# Prerrequisites

Install the following packages:


```r
install.packages("sf")
install.packages("terra")
install.packages("spData")
install.packages(
    "spDataLarge", 
    repos = "https://geocompr.r-universe.dev")
```

For linux check the [Posit Public Package Manager](https://packagemanager.posit.co) and apply the `SETUP` for your case. This repository is configured by default using Positron. Check:

```r
getOption("repos")
```

For example for `Ubuntu 24.04 LTS`:

```
                                                          P3M 
"https://packagemanager.posit.co/cran/__linux__/noble/latest" 
                                                         CRAN 
                                "https://cloud.r-project.org" 
```

# Introduction

Fundamental geographic data models:

- Vector
- Raster

The **vector data model** represents the world using points, lines and polygons. The **raster data model** divides the surface up into cells of constant size.

Which to use?

- Vector data tends to dominate the social sciences because human settlements tend to have discrete borders
- Raster dominates many enfironmental sciences partially becasue the reliance on remote sensing data

# Vector data

The geographic vecto data model is based on points located withing a coordinate reference system, **CRS**.

Points ca represent self-standing features or they can be linked together to form more complex geometries such as lines and polygons.

IN this system, for example, London can be represented by the coordinates `c(-0.1, 51.5)`. This means that its location is `-0.1` degrees east and `51.5` degrees north of the origin. The origin in this case is at `0` degrees longitude (Prime Meridian) and `0` degrees latitude (Equator) in a geographic (lon/lat) CRS.

For this section, it is sufficient to know that coordinates consist of two numbers representing distances from an origin, usually in $x$ and $y$ dimensions.

## `sf` package

- **GDAL**: reading, writing and manipulating geographic data formats
- **PROJ**: coordinate system transformations
- **GEOS**: planar geometry engine for operations such as calculating buffers and centroids on data with a projected CRS
- **S2**: a spherical geometry engine

As an example when loading `sf`:

```
Linking to GEOS 3.12.1, GDAL 3.8.4, PROJ 9.4.0; sf_use_s2() is TRUE
```

### Simple features

Simple features is an open standard developed and endorsed by the Open Geospatial Consortium (OGC). Simple features is a hierarchical data model taht represents a wide rande of geometry types.

Simple feature types fully supported by `sf`:

- Multilinestring
- Multipoint
- Multipolygon
- Geometry collection
- Polygon
- Point
- Line string

Raster data classes are not supported by `sf`. For more information check the site [Simple Features for R](https://r-spatial.github.io/sf/) and the vignettes:

```r
vignette(package = "sf")
vignetter("sf1")
```

Simple feature objects in R are store in a data frame, with geographic data occupying a special column, usually name `geom` or `geometry`.

We are going to use the `world` dataset provided by `spData`:

```{r}
class(world)
names(world)
```

`sf` objects can be plotted quickly with the function `plot()`. For example:

```{r}
plot(world, max.plot = 10)
```

In this case 10 variables from `world` are plot except `geom` where are the polygons. Also treating geographic objects as regular data frames with spatial powers nas many advantages, especially if you are already used to working with data frames.

```{r}
world["lifeExp"] |>
  summary()
```

In this case also the geometry (`geom`) is reported and the world `MULTIPOLYGON` in the output referts to the geometry type of features (countries) in the `world` object.

`sf` objects are easy to subset. For example you can return only the first 2 rows and the first 3 columns of the `world` object.

```{r}
world_mini <- world[1:2, 1:3]
world_mini
```

In a `sf` object, the geometry column is considered the defining characteristic of the data. Without the geometry, the object ceases to be spatial. That is why, `sf` is designed to ensure that the spatial integrity of the features remains intact.

The output shows two major differences compared with a regular `data.frame`: the inclusion of additional geographic metadata (`Geometry type`, `Dimension`, `Bounding box` and coordinate references system information). Also it has the `geom` column.

The `geom` column are list columns of class `sfg`:

```{r}
world$geom |>
  class()
```

`sfc` objects are composed of one or more objects of class `sfg`: simple features geometries.

### Why simple features?

- Fast reading and writing of data
- Enhanced plotting performance
- `sf` objects can be treated as data frames in most operations
- `sf` function names are relatively consistent and intuitive (all begin with `st_`)
- `sf` functions can be combined with the `|>` operator and works well with the `tidyverse` collection of R packages

`sf` support for the `tidyverse` packages is exemplified by `read_sf()`, a function for importing geographic vector data. Unlike the function `st_read()`, which returns attributes stored in a base R `data.frame`, `read_sf()` silently returns data as a tidyverse `tibble`. For example:

```{r}
world_dfr <- system.file(
  "shapes/world.gpkg",
  package = "spData"
) |>
  st_read()

world_tbl <- system.file(
  "shapes/world.gpkg",
  package = "spData"
) |>
  read_sf()

class(world_dfr)
class(world_tbl)
```

`world_tbl` include also the classes `tbl_df` and `tbl` additionally to `sf` and `data.frame`.

### Basic maps

Basic geographic visualizations (maps) are created in `sf` with base R `plot()` function. Be default, this creates a multi-panel plot, one sub-plot for each variable of the object. For example:

```{r}
plot(world[3:6])
plot(world["pop"])
```

`plot()` is a generic method that behaves differently depending on the class of object beign plotted. For example, you can check `methods(plot)`:

```{r}
methods(plot)[methods(plot) == "plot.sf"]
```

Plots are added as layers to existing images by setting `add = TRUE`. For example lets filter the countries in Asia and combines then into a single feature:

```{r}
world_asia <- world[world$continent == "Asia", ]
# A single MULTIPOLYGON is created that represents Asia
# as a single unit
asia <- st_union(world_asia)
plot(asia)
```

We can now plot the Asian continent over a map of the world. The first plot must only have one facet for `add = TRUE`. Also if the first plot has a key (a legend, for example a color scale to represent the data) then you need to specify `reset = FALSE`

```{r}
plot(world["pop"], reset = FALSE)
plot(asia, add = TRUE, col = "red")
```

**Note**: This plot is problematic because the legend shows a fill scale to represent the population for the countries except asia.

Lets try another visualization by adding circles that represent the population on a map of the world as an unprojected version:

```{r}
plot(world["continent"], reset = FALSE)
cex <- sqrt(world$pop) / 10000
world_cents <- st_centroid(world, of_largest = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)
```