---
title: Attribute data operations
format: 
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
---

```{r}
#| label: load-r-packages
#| message: false
library(sf)
library(terra)
library(tidyverse)
library(spData)
```

# Vector Attribute manipulation

Geographic vector datasets are well supported in R thanks to the `sf` class, which extends base R's `data.frame`.

`sf` objects differ from basic data frames because they have a `geometry` column of class `sfc` which contain a range of geographic entities

**sf** also provides generics that allow `sf` objects to behavie like regular data frames, as shown by printing the class's methods:

```{r}
methods(class = "sf")[1:10]
```

Let's start by using base R functions to learn about the `world` dataset from the `spData` package:

```{r}
class(world)
```

```{r}
dim(world)
```

Dropping geometry column before working with attribute data can be useful. Data manipulation processes can run faster when they work only on the attribute data and geometry columns are not always needed.

```{r}
world_df <- st_drop_geometry(world)
ncol(world_df)
```

## Vector attribute subsetting

Base R subsettings methods include the operator `[` and the function `subset()`. The key `dplyr` subsetting functions are `filter()` and `slice()` for subsetting rows and `select()` for subsetting columns. Both approaches preserve the spatial components of attribute data in `sf` objetcs.

Using the operation `$` or the `dplyr` function `pull()` will lose the geometry data.

The command `object[i,j]` means return the rows represented by `i` and the columns represented by `j`, where `i` and `j` typically contain integers or `TRUE`'s and `FALSE`'s.

```{r}
# Subset rows by position
world[1:6, ]
# Subset columns by position
world[, 1:3]
# Subset rows and columns by position
world[1:6, 1:3]
# Columns by name
world[, c("name_long", "pop")]
# Logical indices
world[, c(
  TRUE,
  TRUE,
  FALSE,
  FALSE,
  FALSE,
  FALSE,
  FALSE,
  TRUE,
  TRUE,
  FALSE,
  FALSE
)]
```

A demostration of the utility of using `logical` vectors for subsetting is shown in the code chunk below. This creates a new object, `small_countries`, containing nations whose surface area is smallerthat 10,000 km^2.

```{r}
i_small = world$area_km2 < 10000
summary(i_small)
small_countries = world[i_small, ]
small_countries
```

A more concise command, which omits the intermediary object, generates the same result:

```{r}
small_countries = world[world$area_km2 < 10000, ]
small_countries
```

The base R function `subset()` provides another way to achieve the same result:

```{r}
small_countries = subset(world, area_km2 < 10000)
small_countries
```

We have the following alternatives to extract variables:

```{r}
pull(world, pop)[1:5]
world$pop[1:5]
world[["pop"]][1:5]
```

## Chaining commands with pipes

The native pipe operator `|>` takes its name from the Unix `|`. Pipes enable expressive code: the output of a previous function becomes the first argument of the next function, enabling chaining.

```{r}
world7 = world |>
  filter(continent == "Asia") |>
  select(name_long, continent) |>
  slice(1:5)
world7
```

Another alternative is to split the operations into multiple self-contained lines, which is recommended when developing new R packages, an approach which has the advantage of saving intermediate results with disting names which can be later inspected for debugging purposes:

```{r}
world9_filtered = filter(world, continent == "Asia")
world9_selected = select(world9_filtered, continent)
world9 = slice(world9_selected, 1:5)
world9
```

## Vector attribute aggregation

Aggregation involves summarizing data with one or more grouping variables, typically from columns in the data frame to be aggregated.

We can use the `world` dataset. The aim is to find the `sum()` of country populations for each continent, resulting in a smaller data frame. This can be done with the base R function `aggregate()` as follows:

```{r}
world_agg1 = aggregate(
  pop ~ continent,
  FUN = sum,
  data = world,
  na.rm = TRUE
)
world_agg1
```

The result is non-spatial data frame with six rows, one per continent, and two columns reporting the name and population of each continent.

`aggregate()` is a generic function which means that it behaves differently depending on its inputs. `sf` provides the method `aggregate.sf()` which is activated automatically when `x` is an `sf` object and a `by` argument is provided:

```{r}
world_agg2 = aggregate(
  world["pop"],
  by = list(world$continent),
  FUN = sum,
  na.rm = TRUE
)
world_agg2
```

Another alternative is:

```{r}
world_agg3 = world |>
  group_by(continent) |>
  summarize(pop = sum(pop, na.rm = TRUE))
world_agg3
```

You can change many functions in the following way:

```{r}
world_agg5 = world |>
  st_drop_geometry() |>
  select(pop, continent, area_km2) |>
  group_by(continent) |>
  summarise(
    pop = sum(pop, na.rm = TRUE),
    area = sum(area_km2),
    n = n()
  ) |>
  ungroup() |>
  mutate(density = round(pop / area)) |>
  slice_max(pop, n = 3) |>
  arrange(desc(n))
world_agg5
```

## Vector attribute joining

Combining data from different sources is a common task in data preparation. Joins do this by combining tables base on a shared "key" variable. `dplyr` has multiple join functions including `lef_join()` and `inner_join()`.

To demonstrate joins, we will combine data on coffe production with the `world` dataset. The coffee data is in a data frame called `coffe_data` from the `spData` package. It has three columns:

- `name_long` names major coffee-producing nations
- `coffee_production_2016` contain estimated values for coffe production in units of 60-kg bags in each 2016 
- `coffee_production_2017` contain estimated values for coffe production in units of 60-kg bags in each 2017

```{r}
coffee_data |>
  glimpse()
```

Applying the join we have:


```{r}
world_coffee = world |>
  left_join(
    y = coffee_data,
    by = join_by(name_long)
  )

class(world_coffee)
nrow(world_coffee)
```

```{r}
names(world_coffee)
```

```{r}
plot(world_coffee[, "coffee_production_2017"])
```

For joining to work, a "key" variable must be supplied in both datasets. By default, `dplyr` uses all variables with matching names. In this cas, both `coffee_data` and `world` objects contained a variable `name_long`.

In the majority of cases where variable names are not the same, you have 2 options:

1. Rename the key variable in one of the objects so they match
2. Use the `by` argumetn to specify the joining variables

In the case of `left_join` all 177 country records are kept intact in `world_coffee`. Rows in the original dataset with no match are assigned `NA` values for the new coffee production variables. What if we only want to keep countries that have a match in the key variable? In that case, an inner join can be used:

```{r}
world_coffee_inner = world |>
  inner_join(
    y = coffee_data,
    by = join_by(name_long)
  )
nrow(world_coffee_inner)
```

## Creating attributes and removing spatial information

Ofte, we would like to create a new column based on already existing columns. For example, we want to calcualte population density for each country. For this we need to divide a population column, here `pop`, by an area column, here `area_km2` with unit are in square kilometers. Using base R, we can type:

```{r}
world_new = world |>
  mutate(pop_dens = pop / area_km2)
world_new |>
  glimpse()
```
